 ---
title: "INFO 6105 Final Project — Analysis 2: One-Way ANOVA"
author: "Tasmiya Parveen | NUID: 002509350"
date: "April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

This analysis investigates whether average flow duration differs significantly across the three network protocol types (tcp, udp, icmp) in the RT-IoT2022 dataset.

**Research Question:** Does average flow duration differ significantly across the three network protocol types (tcp, udp, icmp)?

---

## 2. Load Packages

```{r packages, eval=FALSE}
install.packages("ggplot2",    INSTALL_opts = "--no-lock")
install.packages("dplyr",      INSTALL_opts = "--no-lock")
install.packages("car",        INSTALL_opts = "--no-lock")
install.packages("multcomp",   INSTALL_opts = "--no-lock")
install.packages("effectsize", INSTALL_opts = "--no-lock")
```

```{r load-packages}
library(ggplot2)
library(dplyr)
library(car)
library(multcomp)
library(effectsize)
```

---

## 3. Load and Prepare Data

```{r load-data}
df <- read.csv("data/RT_IOT2022.csv")
```

```{r prepare-data}
df_anova <- df %>%
  filter(
    proto %in% c("tcp", "udp", "icmp"),
    is.finite(flow_duration),
    flow_duration > 0
  ) %>%
  mutate(
    log_duration = log(flow_duration),
    proto        = factor(proto)
  )

cat("Rows for ANOVA:", nrow(df_anova), "\n")
cat("Protocol counts:\n")
print(table(df_anova$proto))
```

**Note on group imbalance:** The icmp group has only 24 observations compared to tcp (99,425) and udp (7,687). This imbalance likely contributed to assumption violations discussed in Section 6.

---

## 4. Exploratory Visualizations

### 4.1 Boxplot of Log(Flow Duration) by Protocol

```{r boxplot, fig.width=7, fig.height=5}
ggplot(df_anova, aes(x = proto, y = log_duration, fill = proto)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.6) +
  scale_fill_manual(values = c(
    "tcp"  = "#7F77DD",
    "udp"  = "#1D9E75",
    "icmp" = "#EF9F27"
  )) +
  labs(
    title = "Log(flow duration) by protocol type",
    x     = "Protocol type",
    y     = "log(flow duration in microseconds)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### 4.2 Group Means with Error Bars

```{r means-plot, fig.width=6, fig.height=5}
summary_stats <- df_anova %>%
  group_by(proto) %>%
  summarise(
    n            = n(),
    mean_log_dur = mean(log_duration),
    sd_log_dur   = sd(log_duration),
    .groups      = "drop"
  )

print(summary_stats)

ggplot(summary_stats, aes(x = proto, y = mean_log_dur, fill = proto)) +
  geom_col(width = 0.5, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = mean_log_dur - sd_log_dur,
        ymax = mean_log_dur + sd_log_dur),
    width = 0.2, linewidth = 0.8
  ) +
  scale_fill_manual(values = c(
    "tcp"  = "#7F77DD",
    "udp"  = "#1D9E75",
    "icmp" = "#EF9F27"
  )) +
  labs(
    title = "Mean log(flow duration) by protocol (± 1 SD)",
    x     = "Protocol type",
    y     = "Mean log(flow duration)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

## 5. One-Way ANOVA

```{r anova}
anova_model <- aov(log_duration ~ proto, data = df_anova)
summary(anova_model)
```

**Interpretation:** The F-statistic of 9,448 with p < 2e-16 indicates that at least one protocol group has a significantly different mean log flow duration. The result is highly statistically significant.

---

## 6. Assumption Checks

### 6.1 Normality — Shapiro-Wilk Test

```{r shapiro}
set.seed(42)
residuals_sample <- sample(residuals(anova_model), 5000)
shapiro_result   <- shapiro.test(residuals_sample)
print(shapiro_result)
```

### 6.2 Equal Variances — Levene's Test

```{r levene}
levene_result <- leveneTest(log_duration ~ proto, data = df_anova)
print(levene_result)
```

**Interpretation of assumption violations:** Both the Shapiro-Wilk test (W = 0.533, p < 0.001) and Levene's test (F = 14.54, p < 0.001) indicate violations of normality and equal variance assumptions. However, given the large sample size (n = 107,136), ANOVA is considered robust to these violations. The extreme imbalance in the icmp group (n = 24) likely contributed to both violations and is noted as a limitation.

---

## 7. Tukey HSD Post-Hoc Test

```{r tukey}
tukey_result <- TukeyHSD(anova_model)
print(tukey_result)
```

**Interpretation:** All three pairwise comparisons are statistically significant. The order of mean log flow duration from longest to shortest is icmp > udp > tcp. This makes practical sense — ICMP is used for diagnostic pings that can remain active longer, while TCP connections in attack traffic tend to be very short bursts.

---

## 8. Effect Size — Eta Squared (η²)

```{r effect-size}
eta_result <- eta_squared(anova_model)
print(eta_result)
```

**Interpretation:** η² = 0.15 indicates a large effect size. Protocol type explains 15% of the variation in flow duration, meaning protocol type is not only statistically significant but also practically meaningful.

---

## 9. Answer to Research Question

The one-way ANOVA reveals that average flow duration differs significantly across the three protocol types (F = 9,448, p < 2e-16). Tukey HSD post-hoc tests confirm that all three pairs — tcp vs icmp, udp vs icmp, and udp vs tcp — differ significantly. The effect size is large (η² = 0.15), indicating that protocol type is a practically meaningful predictor of flow duration in IoT network traffic.

---

## 10. Session Info

```{r session-info}
sessionInfo()
```