 ---
title: "INFO 6105 Final Project — Analysis 3: Logistic Regression"
author: "Tasmiya Parveen | NUID: 002509350"
date: "April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

This analysis investigates whether flow-level characteristics can predict whether an IoT network connection is an intrusion or normal traffic using the RT-IoT2022 dataset.

**Research Question:** Can flow-level characteristics (flow duration, forward packet rate, backward packet rate, average payload size) predict whether an IoT network connection is an intrusion or normal traffic?

---

## 2. Load Packages

```{r packages, eval=FALSE}
install.packages("ggplot2", INSTALL_opts = "--no-lock")
install.packages("dplyr",   INSTALL_opts = "--no-lock")
install.packages("pROC",    INSTALL_opts = "--no-lock")
install.packages("caret",   INSTALL_opts = "--no-lock")
```

```{r load-packages}
library(ggplot2)
library(dplyr)
library(pROC)
library(caret)
```

---

## 3. Load and Prepare Data

```{r load-data}
df <- read.csv("data/RT_IOT2022.csv")
```

```{r prepare-data}
df_logistic <- df %>%
  filter(
    is.finite(flow_duration)         & flow_duration > 0,
    is.finite(fwd_pkts_per_sec),
    is.finite(bwd_pkts_per_sec),
    is.finite(flow_pkts_payload.avg) & flow_pkts_payload.avg > 0
  ) %>%
  mutate(
    intrusion    = ifelse(Attack_type %in% c("MQTT_Publish",
                                              "Thing_Speak",
                                              "Wipro_bulb"), 0, 1),
    log_duration = log(flow_duration),
    log_payload  = log(flow_pkts_payload.avg)
  )

cat("Total rows:", nrow(df_logistic), "\n")
cat("Class distribution:\n")
print(table(df_logistic$intrusion))
cat("Normal %:", round((1 - mean(df_logistic$intrusion)) * 100, 2), "%\n")
cat("Attack %:", round(mean(df_logistic$intrusion) * 100, 2), "%\n")
```

**Note on class imbalance:** The derived binary variable has approximately 88% attack and 12% normal traffic. This reflects the nature of the dataset which was collected during active attack simulations. Model evaluation will use precision, recall, F1-score, and AUC in addition to accuracy to account for this imbalance.

---

## 4. Exploratory Visualizations

### 4.1 Intrusion Rate by Protocol Type

```{r intrusion-rate-plot, fig.width=7, fig.height=5}
df_proto_plot <- df_logistic %>%
  filter(proto %in% c("tcp", "udp", "icmp")) %>%
  group_by(proto) %>%
  summarise(intrusion_rate = mean(intrusion), .groups = "drop")

ggplot(df_proto_plot, aes(x = proto, y = intrusion_rate, fill = proto)) +
  geom_col(width = 0.5, alpha = 0.85) +
  scale_fill_manual(values = c(
    "tcp"  = "#7F77DD",
    "udp"  = "#1D9E75",
    "icmp" = "#EF9F27"
  )) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Intrusion rate by protocol type",
    x     = "Protocol type",
    y     = "Proportion of attack traffic"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### 4.2 Forward Packets/Sec by Intrusion Status

```{r density-plot, fig.width=7, fig.height=5}
df_density <- df_logistic %>%
  filter(fwd_pkts_per_sec < quantile(fwd_pkts_per_sec, 0.99)) %>%
  mutate(intrusion_label = ifelse(intrusion == 1, "Attack", "Normal"))

ggplot(df_density, aes(x = fwd_pkts_per_sec, fill = intrusion_label)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Attack" = "#D85A30", "Normal" = "#1D9E75")) +
  labs(
    title = "Forward packets/sec by intrusion status",
    x     = "Forward packets per second",
    y     = "Density",
    fill  = "Traffic type"
  ) +
  theme_minimal(base_size = 13)
```

---

## 5. Train/Test Split

```{r train-test-split}
set.seed(42)
n         <- nrow(df_logistic)
train_idx <- sample(1:n, size = 0.7 * n)
df_train  <- df_logistic[train_idx, ]
df_test   <- df_logistic[-train_idx, ]

cat("Training rows:", nrow(df_train), "\n")
cat("Testing rows:", nrow(df_test), "\n")
```

---

## 6. Fit Logistic Regression Model

```{r fit-model, warning=FALSE}
logit_model <- glm(
  intrusion ~ log_duration + fwd_pkts_per_sec +
              bwd_pkts_per_sec + log_payload,
  data   = df_train,
  family = binomial
)

summary(logit_model)
```

**Note:** The warning "fitted probabilities numerically 0 or 1 occurred" indicates complete separation — our predictors are so strongly associated with the outcome that some predicted probabilities are essentially certain. This is a sign of strong predictors rather than a modeling error.

---

## 7. Odds Ratios

```{r odds-ratios}
odds_ratios <- exp(coef(logit_model))
conf_int    <- exp(coef(logit_model) +
               outer(sqrt(diag(vcov(logit_model))), c(-1.96, 1.96)))
colnames(conf_int) <- c("Lower 95%", "Upper 95%")

cat("Odds Ratios with 95% Confidence Intervals:\n")
print(cbind(OR = round(odds_ratios, 4), round(conf_int, 4)))
```

**Interpretation (holding all other variables constant):**

- **`log_duration`** (OR = 1.21): Each unit increase in log flow duration makes an attack **1.21 times more likely**
- **`fwd_pkts_per_sec`** (OR = 1.004): Each unit increase in forward packet rate slightly increases attack probability
- **`bwd_pkts_per_sec`** (OR = 1.0003): Small but significant increase in attack probability
- **`log_payload`** (OR = 3.18): Each unit increase in log payload makes an attack **3.18 times more likely** — the strongest predictor

---

## 8. Predictions and Confusion Matrix

```{r predictions}
pred_probs  <- predict(logit_model, newdata = df_test, type = "response")
pred_labels <- ifelse(pred_probs > 0.5, 1, 0)
```

```{r confusion-matrix}
conf_matrix <- confusionMatrix(
  factor(pred_labels, levels = c(0, 1)),
  factor(df_test$intrusion, levels = c(0, 1)),
  positive = "1"
)
print(conf_matrix)

cat("\nKey metrics:\n")
cat("Accuracy:   ", round(conf_matrix$overall["Accuracy"], 4), "\n")
cat("Sensitivity:", round(conf_matrix$byClass["Sensitivity"], 4), "\n")
cat("Specificity:", round(conf_matrix$byClass["Specificity"], 4), "\n")
cat("Precision:  ", round(conf_matrix$byClass["Precision"], 4), "\n")
cat("F1 Score:   ", round(conf_matrix$byClass["F1"], 4), "\n")
```

---

## 9. ROC Curve and AUC

```{r roc-curve, fig.width=6, fig.height=5}
roc_obj <- roc(df_test$intrusion, pred_probs)
auc_val <- auc(roc_obj)
cat("AUC:", round(auc_val, 4), "\n")

plot(roc_obj,
     col  = "#7F77DD",
     lwd  = 2,
     main = paste("ROC Curve — AUC =", round(auc_val, 4)),
     print.auc = TRUE)
abline(a = 0, b = 1, lty = 2, col = "gray")
```

**Interpretation:** AUC = 0.9937 indicates near-perfect discriminative ability. The model correctly ranks a random attack above a random normal connection 99.37% of the time.

---

## 10. Threshold Tuning

```{r threshold-tuning}
thresholds <- seq(0.1, 0.9, by = 0.1)
results    <- data.frame()

for (thresh in thresholds) {
  preds <- ifelse(pred_probs > thresh, 1, 0)
  tp    <- sum(preds == 1 & df_test$intrusion == 1)
  fp    <- sum(preds == 1 & df_test$intrusion == 0)
  tn    <- sum(preds == 0 & df_test$intrusion == 0)
  fn    <- sum(preds == 0 & df_test$intrusion == 1)

  accuracy    <- (tp + tn) / (tp + fp + tn + fn)
  sensitivity <- ifelse((tp + fn) > 0, tp / (tp + fn), NA)
  specificity <- ifelse((tn + fp) > 0, tn / (tn + fp), NA)
  f1          <- ifelse((tp + fp + fn) > 0,
                        2 * tp / (2 * tp + fp + fn), NA)

  results <- rbind(results, data.frame(
    threshold   = thresh,
    accuracy    = round(accuracy, 4),
    sensitivity = round(sensitivity, 4),
    specificity = round(specificity, 4),
    f1_score    = round(f1, 4)
  ))
}

print(results)
```

**Threshold choice:** A threshold of 0.3 was selected as optimal because it yields the highest F1 score (0.9848) while maintaining 98.34% sensitivity. In cybersecurity, missing an attack is more costly than a false alarm, so maximizing sensitivity is the priority.

---

## 11. Answer to Research Question

The logistic regression model demonstrates that flow-level characteristics are highly effective predictors of intrusion status. Log payload size is the strongest predictor (OR = 3.18, p < 0.001), followed by forward packet rate and flow duration. The model achieves AUC = 0.9937, accuracy = 96.83%, and F1 = 98.19% at the default threshold. Using an optimized threshold of 0.3 further improves sensitivity to 98.34%, which is critical for real-world intrusion detection where missing an attack has severe consequences.

---

## 12. Session Info

```{r session-info}
sessionInfo()
```