---
title: "INFO 6105 Final Project — Analysis 1: Multiple Linear Regression"
author: "Tasmiya Parveen | NUID: 002509350"
date: "April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

This analysis investigates which flow-level features best predict the average payload size per packet (`flow_pkts_payload.avg`) in IoT network connections using the RT-IoT2022 dataset from the UCI Machine Learning Repository.

**Research Question:** Which flow-level features (forward packet rate, flow duration, and down/up ratio) best predict the average payload size per packet in an IoT network connection?

---

## 2. Load Packages

```{r packages}
library(ggplot2)
library(dplyr)
library(car)
library(GGally)
library(reshape2)
library(broom)
```

---

## 3. Load and Clean Data

```{r load-data}
df <- read.csv("data/RT_IOT2022.csv")

cat("Total rows loaded:", nrow(df), "\n")
cat("Total columns:", ncol(df), "\n")
```

```{r clean-data}
df_clean <- df %>%
  filter(
    is.finite(flow_pkts_payload.avg) & flow_pkts_payload.avg > 0,
    is.finite(fwd_pkts_per_sec),
    is.finite(bwd_pkts_per_sec),
    is.finite(flow_duration)        & flow_duration > 0,
    is.finite(down_up_ratio)
  )

cat("Rows after cleaning:", nrow(df_clean), "\n")

set.seed(42)
df_sample <- df_clean %>% sample_n(10000)

df_sample <- df_sample %>%
  mutate(
    log_payload  = log(flow_pkts_payload.avg),
    log_duration = log(flow_duration)
  )

cat("Sample size:", nrow(df_sample), "\n")
```

---

## 4. Exploratory Visualizations

### 4.1 Scatterplot Matrix

```{r scatterplot-matrix, fig.width=9, fig.height=8}
plot1_data <- df_sample %>%
  select(log_payload, fwd_pkts_per_sec,
         bwd_pkts_per_sec, log_duration, down_up_ratio)

ggpairs(
  plot1_data,
  title = "Scatterplot matrix: predictors vs log(payload avg)",
  upper = list(continuous = wrap("cor", size = 3.5)),
  lower = list(continuous = wrap("points", alpha = 0.15, size = 0.4)),
  diag  = list(continuous = wrap("densityDiag", fill = "#7F77DD", alpha = 0.5))
) +
  theme_minimal(base_size = 11)
```

### 4.2 Correlation Heatmap

```{r correlation-heatmap, fig.width=7, fig.height=6}
cor_matrix <- cor(plot1_data, use = "complete.obs")
cor_melted  <- melt(cor_matrix)

ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3.5) +
  scale_fill_gradient2(
    low = "#3B8BD4", mid = "white", high = "#D85A30",
    midpoint = 0, limits = c(-1, 1), name = "Correlation"
  ) +
  labs(
    title = "Correlation heatmap of regression variables",
    x = "", y = ""
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## 5. Initial Model — All 4 Predictors

```{r initial-model}
model <- lm(log_payload ~ fwd_pkts_per_sec + bwd_pkts_per_sec +
              log_duration + down_up_ratio,
            data = df_sample)

summary(model)
```

### 5.1 VIF Check — Multicollinearity

```{r vif-initial}
cat("VIF values for initial model:\n")
vif(model)
```

**Interpretation:** `fwd_pkts_per_sec` and `bwd_pkts_per_sec` have VIF values exceeding 200,000 — indicating extreme multicollinearity. Both variables are derived from the same underlying packet flow. `bwd_pkts_per_sec` is dropped and the model is refit.

---

## 6. Final Model — After Removing Multicollinear Predictor

```{r final-model}
model2 <- lm(log_payload ~ fwd_pkts_per_sec + 
               log_duration + down_up_ratio,
             data = df_sample)

summary(model2)
```

### 6.1 VIF Check — Final Model

```{r vif-final}
cat("VIF values for final model:\n")
vif(model2)
```

All VIF values are below 5 — no multicollinearity issues in the final model. ✅

---

## 7. Coefficient Interpretation

```{r coefficients}
tidy(model2)
```

**Holding all else constant:**

- **`fwd_pkts_per_sec`** (-5.39e-08, p < 0.001): For each unit increase in forward packet rate, log(payload avg) decreases by 5.39e-08. Statistically significant — high packet rates (typical of flood attacks) are associated with smaller payloads.
- **`log_duration`** (-2.27e-03, p = 0.083): For each unit increase in log(flow duration), log(payload avg) decreases slightly. Not significant at the 0.05 level — flow duration is a weak predictor of payload size.
- **`down_up_ratio`** (2.199, p < 0.001): For each unit increase in the down/up ratio, log(payload avg) increases by 2.199. The strongest predictor — connections with more download than upload traffic carry larger payloads.

---

## 8. Model Fit

```{r model-fit}
r_squared     <- summary(model2)$r.squared
adj_r_squared <- summary(model2)$adj.r.squared
f_stat        <- summary(model2)$fstatistic

cat("R-squared:         ", round(r_squared, 4), "\n")
cat("Adjusted R-squared:", round(adj_r_squared, 4), "\n")
cat("F-statistic:       ", round(f_stat[1], 2), "\n")
cat("Overall p-value:   ", pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = FALSE), "\n")
```

**Interpretation:** The model explains approximately 29% of the variation in log(payload avg) (R² = 0.29). The overall F-test is highly significant (F = 1360.8, p ≈ 0), confirming the model is useful.

---

## 9. Diagnostic Plots

```{r diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(model2)
```

**Interpretation:**

- **Residuals vs Fitted:** Some curvature present — mild non-linearity in the data.
- **Q-Q Plot:** Tails deviate from the line — slight non-normality in residuals, common with large real-world datasets.
- **Scale-Location:** Slight fanning pattern — mild heteroscedasticity present, will be noted as a limitation.

---

## 10. Answer to Research Question

The multiple linear regression model reveals that **down/up ratio** is the strongest predictor of average payload size (β = 2.199, p < 0.001), followed by **forward packet rate** (β = -5.39e-08, p < 0.001). Flow duration was not a significant predictor (p = 0.083). The model explains 29% of the variation in payload size (R² = 0.29), which is reasonable for real-world IoT network data with many unmeasured factors.

---

## 11. Session Info

```{r session-info}
sessionInfo()
```